---
title: "new_model"
author: "kaiyu"
date: "2023-04-14"
output: html_document
---
### Introduction
In this file, we will use a decision tree model to predict housing prices in Seattle, Washington in 2014. The data we will use is from Kaggle, uploaded by SHREE. Considering the data provided, we are wrangling a large set of property sales records stored in an unknown format and with unknown data quality issues. Additionally, we will model the data using random forest and cubic spline. The full dataset contains 4600 houses throughout the state of Washington with 18 variables, 1573 of the houses belonging to Seattle. The response variable is price. 

```{r}
#importing dataset
data <- read.csv('~/Desktop/math_448/Project/House_Price_Prediction/data.csv')
summary(data)
str(data)
```
##dropping unnecessary variables(view, date, street, statezip, country)
```{r}
data$rooms <- data$bedrooms + data$bathrooms
data <- subset(data, select = -c(view, date, street, statezip, country, bedrooms, bathrooms))
# Add a new column 'total_rooms' to the dataset
data
```


##spliting the dataset
```{r}
library(caTools)
library(rpart)
set.seed(123) 
split = sample.split(data$price, SplitRatio = 0.7)
training_set = subset(data, split == TRUE)
test_set = subset(data, split == FALSE)
test_set
```

##fitting a tree regression model
```{r}
regressor = rpart(price ~ ., data = training_set, control = rpart.control(minsplit = 20))
summary(regressor)
```

##fitting a test
```{r}
y_pred = predict(regressor, newdata = test_set)
```

##using RMSE to see how much it fits
```{r}
rmse = sqrt(mean((test_set$price - y_pred)^2))
print(paste('RMSE:', rmse))
```
##Visualize the fit
```{r}
# Predict values on your test data
predictions <- predict(regressor, newdata = test_set)

# Plot predicted vs actual values
plot(test_set$price, predictions, main="Predicted vs Actual Values",
     xlab="Actual", ylab="Predicted", xlim=range(0, max(test_set$price)),
     ylim=range(0, max(test_set$price)))
abline(a = 0, b = 1, col = "red")  # adds a y=x line indicating perfect prediction
```



##Visualizing the fit of a decision tree regression
```{r}
library(rpart.plot)
rpart.plot(regressor)
rpart.plot(regressor, compress = TRUE, fallen.leaves = TRUE)
```



