---
title: "project X"
output: html_document
date: "2023-03-30"
---
vewvegweg

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data <- read.csv('~/Desktop/math_448/Project/House_price_prediction/data.csv')
print(paste("Number of records: ", nrow(data)))
print(paste("Number of features: ", ncol(data)))
head(data)
```


```{r}
summary(data)
colnames(data)
str(data)
```
####Adjusting dataset

#checking variables
```{r}
length(unique(data$city)) #There are 44 unique city. 
length(unique(data$country)) #There is only 1 country which should mean nothing
length(unique(data$statezip)) #77 different kinds of statezip
length(unique(data$street)) #4525 that is too much

```

#From this result
we are going to drop "street" and "country" variables because one country means no effect on prediction and 4525 country seems to be too much to include.

We will convert "city" and "state" into categorical variables. 
```{r}
new_data = subset(data, select = -c(street, country) )
head(new_data)
new_data$city<-as.factor(new_data$city)
new_data$statezip<-as.factor(new_data$statezip)
str(new_data)
```
```{r}

#I want to change the code here to look better. 

library("stringr")   
tail(new_data$date, n=100)
str_count(new_data$date, "2014")
```
It looks like all the housings are from the year of 2014.Therefore we are just going to drop this column and make a new column that indicate the years of the house. To show that, we are going to subtract "2014" by "the years it was built". 

```{r}
new_data = subset(new_data, select = -c(date))
sum(is.na(new_data))
head(new_data)
```

```{r}
fit <- lm(price ~ bedrooms + bathrooms + sqft_living + sqft_lot + floors + waterfront + view + condition + sqft_above + sqft_basement + yr_built + yr_renovated + city + statezip, data = new_data)
summary(fit)
```
##Linear Regression with all variables look quite bad model since R squared is around 0.3. 
Let's drop two categorical columns "statezip" and "city"

```{r}
new_fit <- lm(price ~ bedrooms + bathrooms + sqft_living + sqft_lot + floors + waterfront + view + condition + sqft_above + sqft_basement + yr_built + yr_renovated, data = new_data)
summary(new_fit)
```
##The model seems to be worse than before.
```{r}
```


```{r}
a = dbinom(14, size=15, prob=0.67) + dbinom(15, size=15, prob=0.67) 
a
```







